### 2018-07-05 submission/manuscript.Rmd
update first part of results with gc

***Human and mock samples:***
Fecal samples were obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN). The processing and storage of these samples were previously published [@erin_seekatz_2016]. Other than confirmation that none of these individuals had a *Clostridium difficle* infection, clinical data and other types of meta data were not utilized or accessed for this study. All samples were extracted using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA). The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used for mock communities and was made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).

***PCR protocol:***
The five different high fidelity DNA polymerases (hereto referred to as polymerases) that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The polymerases activation time was 2 minutes, unless a different activation was specified by the manufacturer. The annealing and extension time for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md). For Kappa and Q5, the annealing and extension time also were from a previously published protocol [@taq_Gohl_2016]. For Phusion, the company defined activation and annealing times were used while the extension time followed the Accuprime and Platinum settings.

@@

***Sequence processing:***
The mothur software program was used for all sequence processing steps [@mothur_schloss_2009]. The protocol has been previously published [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP). Two major differences from the published protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for OTU generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017]. Sequence error was determined using the 'seq.error' command on mock samples to compare back to the reference 16S sequences of *P. aeruginosa*, *E. coli*, *S. enterica*, *L. fermentum*, *E. faecalis*, *S. aureus*, *L. monocytogenes*, and *B. subtilis* [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].

***Analysis workflow:***
To adjust for unequal sequencing, all samples were rarefied to 1000 sequences for downstream analysis. Analysis of the mock community of each polymerase for GC-based amplification differences, sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras before and after chimera removal with VSEARCH was assessed. Additionally, the correlation between the number of chimeras and the number of OTUs was also assessed. The total number of OTUs, taxa differences, and Bray-Curtis indices were analyzed for both the fecal and mock community samples. Finally, Random Forest models were created to assess whether classification of polymerase, cycles, or individual could be performed best using 16S rRNA gene sequencing data. Additionally, overlap between the most important OTUs to the three models was assessed using mean decrease in accuracy (MDA).

***Statistical analysis:***
All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. Data transformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017]. High and low GC content was determined based on the median GC percentage of either the V4 region or the whole genome of the bacterial species used in the mock community. Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test. For the comparison of the number of OTUS in fecal samples the data was normalized to each individual by cycle number to account for the biological variation. Bray-Curtis distance matrices were generated using mothur. The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and Kruskal-Wallis tests within R. The Random Forest models were run using the caret package (v `r paste(packageVersion("caret")[1])`) [@caret_citation]. A total of 100 10-fold CV runs on different 80/20 splits of the data was run to generate a range of the Logloss value. The probability of a correct call was obtained from this Logloss value by taking the negative natural logrithim. For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test. Where applicable, correction for multiple comparison utilized the Benjamini-Hochberg method [@benjamini_controlling_1995].

***Reproducible methods:***
The code and analysis can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017. The raw sequences can be found on the SRA at the following accession number SRP132931.

@@

***16S rRNA gene sequencing counts differ by GC content:*** There was a significant difference in relative abundance between high/low GC content based on either the V4 16S rRNA gene region or the whole genome [Figure 1, Table S1 & S2].

***Sequence error varies by polymerase and is consistent across the number of cycles used:***
The median error rate varied by polymerase, with Kappa having the highest error rate of all the polymerases across the number of cycles [Figure 4 & Table S3]. The majority of the differences across the number of cycles was between Kappa and the other polymerases [Figure 4 & Table S4]. The total sequences with at least one error was also polymerase dependent, with the majority of differences being between Kappa or Accuprime and the other polymerases [Table S3 & S4]. These differences in error rates were not due to polymerase dependent differences in base substitution rate [Figure S1]. Collectively, the results suggest that sequence error is dependent on polymerase, persists across the number of cycles used, and are not due to any bias towards a specific base substitution.

>>>>

***Human and mock samples.***
Fecal samples were obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN). The processing and storage of these samples were previously published [@erin_seekatz_2016]. Other than confirmation that none of these individuals had a *Clostridium difficle* infection, clinical data and other types of meta data were not utilized or accessed for this study. All samples were extracted using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA). The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used for mock communities and was made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).

***PCR protocol.***
The five different high fidelity DNA polymerases (hereto referred to as polymerases) that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The polymerases activation time was 2 minutes, unless a different activation was specified by the manufacturer. The annealing and extension time for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md). For Kappa and Q5, the annealing and extension time also were from a previously published protocol [@taq_Gohl_2016]. For Phusion, the company defined activation and annealing times were used while the extension time followed the Accuprime and Platinum settings.

@@

***Sequence processing.***
The mothur software program was used for all sequence processing steps [@mothur_schloss_2009]. The protocol has been previously published [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP). Two major differences from the published protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for OTU generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017]. Sequence error was determined using the 'seq.error' command on mock samples to compare back to the reference 16S sequences of *P. aeruginosa*, *E. coli*, *S. enterica*, *L. fermentum*, *E. faecalis*, *S. aureus*, *L. monocytogenes*, and *B. subtilis* [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].

***Analysis workflow.***
To adjust for unequal sequencing, all samples were rarefied to 1000 sequences for downstream analysis. Analysis of the mock community of each polymerase for GC-based amplification differences, sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras before and after chimera removal with VSEARCH was assessed. Additionally, the correlation between the number of chimeras and the number of OTUs was also assessed. The total number of OTUs, taxa differences, and Bray-Curtis indices were analyzed for both the fecal and mock community samples. Finally, Random Forest models were created to assess whether classification of polymerase, cycles, or individual could be performed best using 16S rRNA gene sequencing data. Additionally, overlap between the most important OTUs to the three models was assessed using mean decrease in accuracy (MDA).

***Statistical analysis.***
All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. Data transformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017]. High and low GC content was determined based on the median GC percentage of either the V4 region or the whole genome of the bacterial species used in the mock community. Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test. For the comparison of the number of OTUS in fecal samples the data was normalized to each individual by cycle number to account for the biological variation. Bray-Curtis distance matrices were generated using mothur. The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and Kruskal-Wallis tests within R. The Random Forest models were run using the caret package (v `r paste(packageVersion("caret")[1])`) [@caret_citation]. A total of 100 10-fold CV runs on different 80/20 splits of the data was run to generate a range of the Logloss value. The probability of a correct call was obtained from this Logloss value by taking the negative natural logrithim. For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test. Where applicable, correction for multiple comparison utilized the Benjamini-Hochberg method [@benjamini_controlling_1995].

***Reproducible methods.***
The code and analysis can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017. The raw sequences can be found on the SRA at the following accession number SRP132931.

@@

***Observed relative abundance differs by GC content.***
There was a significant difference in relative abundance between high/low GC content based on either the V4 16S rRNA gene region or the whole genome [Figure 1 & Table S1].

***Sequence error varies by polymerase and is consistent across the number of cycles used.***
The median error rate varied by polymerase, with Kappa having the highest error rate of all the polymerases across the number of cycles [Figure 4 & Table S3]. The majority of the differences across the number of cycles was between Kappa and the other polymerases [Figure 4 & Table S4]. The total sequences with at least one error was also polymerase dependent, with the majority of differences being between Kappa or Accuprime and the other polymerases [Table S3 & S4]. These differences in error rates were not due to polymerase dependent differences in base substitution rate [Figure S1]. Collectively, the results suggest that sequence error is dependent on polymerase, persists across the number of cycles used, and are not due to any bias towards a specific base substitution.

### 2018-07-05 submission/manuscript.tex
update first part of results with gc

\textbf{\emph{Human and mock samples:}} Fecal samples were obtained from
4 individuals who were part of the Enterics Research Investigational
Network (ERIN). The processing and storage of these samples were
previously published (Seekatz et al., 2016). Other than confirmation
that none of these individuals had a \emph{Clostridium difficle}
infection, clinical data and other types of meta data were not utilized
or accessed for this study. All samples were extracted using the
MOBIO\textsuperscript{TM} PowerMag Microbiome RNA/DNA extraction kit
(now Qiagen, MD, USA). The ZymoBIOMICS\textsuperscript{TM} Microbial
Community DNA Standard (Zymo, CA, USA) was used for mock communities and
was made up of \emph{Pseudomonas aeruginosa}, \emph{Escherichia coli},
\emph{Salmonella enterica}, \emph{Lactobacillus fermentum},
\emph{Enterococcus faecalis}, \emph{Staphylococcus aureus},
\emph{Listeria monocytogenes}, and \emph{Bacillus subtilis} at equal
genomic DNA abundance
(\url{http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards}).

\textbf{\emph{PCR protocol:}} The five different high fidelity DNA
polymerases (hereto referred to as polymerases) that were tested
included AccuPrime\textsuperscript{TM} (ThermoFisher, MA, USA), KAPA
HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum
(ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The
polymerases activation time was 2 minutes, unless a different activation
was specified by the manufacturer. The annealing and extension time for
Platinum and Accuprime followed a previously published protocol (Kozich
et al., 2013)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
For Kappa and Q5, the annealing and extension time also were from a
previously published protocol (Gohl et al., 2016). For Phusion, the
company defined activation and annealing times were used while the
extension time followed the Accuprime and Platinum settings.

@@

\textbf{\emph{Sequence processing:}} The mothur software program was
used for all sequence processing steps (Schloss et al., 2009). The
protocol has been previously published (Kozich et al., 2013)
(\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two major differences
from the published protocol were the use of VSEARCH instead of UCHIME
for chimera detection and the use of the OptiClust algorithm instead of
average neighbor for OTU generation at 97\% similarity (Edgar et al.,
2011; Rognes et al., 2016; Westcott \& Schloss, 2017). Sequence error
was determined using the `seq.error' command on mock samples to compare
back to the reference 16S sequences of \emph{P. aeruginosa}, \emph{E.
coli}, \emph{S. enterica}, \emph{L. fermentum}, \emph{E. faecalis},
\emph{S. aureus}, \emph{L. monocytogenes}, and \emph{B. subtilis}
(Schloss et al., 2009; Cole et al., 2013; Rognes et al., 2016).

\textbf{\emph{Analysis workflow:}} To adjust for unequal sequencing, all
samples were rarefied to 1000 sequences for downstream analysis.
Analysis of the mock community of each polymerase for GC-based
amplification differences, sequence error rate, number of sequences with
an error, base substitution, and numbers of chimeras before and after
chimera removal with VSEARCH was assessed. Additionally, the correlation
between the number of chimeras and the number of OTUs was also assessed.
The total number of OTUs, taxa differences, and Bray-Curtis indices were
analyzed for both the fecal and mock community samples. Finally, Random
Forest models were created to assess whether classification of
polymerase, cycles, or individual could be performed best using 16S rRNA
gene sequencing data. Additionally, overlap between the most important
OTUs to the three models was assessed using mean decrease in accuracy
(MDA).

\textbf{\emph{Statistical analysis:}} All analysis was done with the R
(v 3.4.4) software package (R Core Team, 2017). Data transformation and
graphing was completed using the tidyverse package (v 1.2.1) and colors
selected using the viridis package (v 0.4.1) (Garnier, 2017; Wickham,
2017). High and low GC content was determined based on the median GC
percentage of either the V4 region or the whole genome of the bacterial
species used in the mock community. Differences in the total number of
OTUs were analyzed using an ANOVA with a tukey post-hoc test. For the
comparison of the number of OTUS in fecal samples the data was
normalized to each individual by cycle number to account for the
biological variation. Bray-Curtis distance matrices were generated using
mothur. The distance matrix data was analyzed using PERMANOVA with the
vegan package (v 2.4.5) (Oksanen et al., 2017) and Kruskal-Wallis tests
within R. The Random Forest models were run using the caret package (v
6.0.78) (Jed Wing et al., 2017). A total of 100 10-fold CV runs on
different 80/20 splits of the data was run to generate a range of the
Logloss value. The probability of a correct call was obtained from this
Logloss value by taking the negative natural logrithim. For both error
and chimera analysis, samples were tested using Kruskal-Wallis with a
Dunns post-hoc test. Where applicable, correction for multiple
comparison utilized the Benjamini-Hochberg method (Benjamini \&
Hochberg, 1995).

\textbf{\emph{Reproducible methods:}} The code and analysis can be found
here \url{https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017}.
The raw sequences can be found on the SRA at the following accession
number SRP132931.

@@

***16S rRNA gene sequencing counts do not differ by GC \url{content:***}
Although the 16S rRNA gene sequencing counts varied by polymerase used,
there was no difference between V4 16S rRNA genes with high and low GC
content at every amplification cycle {[}Figure 1 \& Table S1{]}.
Additionally, there was also no difference in 16S rRNA gene sequencing
counts between high and low GC content genomes {[}Table S2{]}.

\textbf{\emph{Sequence error varies by polymerase and is consistent
across the number of cycles used:}} The median error rate varied by
polymerase, with Kappa having the highest error rate of all the
polymerases across the number of cycles {[}Figure 4 \& Table S3{]}. The
majority of the differences across the number of cycles was between
Kappa and the other polymerases {[}Figure 4 \& Table S4{]}. The total
sequences with at least one error was also polymerase dependent, with
the majority of differences being between Kappa or Accuprime and the
other polymerases {[}Table S3 \& S4{]}. These differences in error rates
were not due to polymerase dependent differences in base substitution
rate {[}Figure S1{]}. Collectively, the results suggest that sequence
error is dependent on polymerase, persists across the number of cycles
used, and are not due to any bias towards a specific base substitution.

>>>>

\textbf{\emph{Human and mock samples.}} Fecal samples were obtained from
4 individuals who were part of the Enterics Research Investigational
Network (ERIN). The processing and storage of these samples were
previously published (Seekatz et al., 2016). Other than confirmation
that none of these individuals had a \emph{Clostridium difficle}
infection, clinical data and other types of meta data were not utilized
or accessed for this study. All samples were extracted using the
MOBIO\textsuperscript{TM} PowerMag Microbiome RNA/DNA extraction kit
(now Qiagen, MD, USA). The ZymoBIOMICS\textsuperscript{TM} Microbial
Community DNA Standard (Zymo, CA, USA) was used for mock communities and
was made up of \emph{Pseudomonas aeruginosa}, \emph{Escherichia coli},
\emph{Salmonella enterica}, \emph{Lactobacillus fermentum},
\emph{Enterococcus faecalis}, \emph{Staphylococcus aureus},
\emph{Listeria monocytogenes}, and \emph{Bacillus subtilis} at equal
genomic DNA abundance
(\url{http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards}).

\textbf{\emph{PCR protocol.}} The five different high fidelity DNA
polymerases (hereto referred to as polymerases) that were tested
included AccuPrime\textsuperscript{TM} (ThermoFisher, MA, USA), KAPA
HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum
(ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The
polymerases activation time was 2 minutes, unless a different activation
was specified by the manufacturer. The annealing and extension time for
Platinum and Accuprime followed a previously published protocol (Kozich
et al., 2013)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
For Kappa and Q5, the annealing and extension time also were from a
previously published protocol (Gohl et al., 2016). For Phusion, the
company defined activation and annealing times were used while the
extension time followed the Accuprime and Platinum settings.

@@

\textbf{\emph{Sequence processing.}} The mothur software program was
used for all sequence processing steps (Schloss et al., 2009). The
protocol has been previously published (Kozich et al., 2013)
(\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two major differences
from the published protocol were the use of VSEARCH instead of UCHIME
for chimera detection and the use of the OptiClust algorithm instead of
average neighbor for OTU generation at 97\% similarity (Edgar et al.,
2011; Rognes et al., 2016; Westcott \& Schloss, 2017). Sequence error
was determined using the `seq.error' command on mock samples to compare
back to the reference 16S sequences of \emph{P. aeruginosa}, \emph{E.
coli}, \emph{S. enterica}, \emph{L. fermentum}, \emph{E. faecalis},
\emph{S. aureus}, \emph{L. monocytogenes}, and \emph{B. subtilis}
(Schloss et al., 2009; Cole et al., 2013; Rognes et al., 2016).

\textbf{\emph{Analysis workflow.}} To adjust for unequal sequencing, all
samples were rarefied to 1000 sequences for downstream analysis.
Analysis of the mock community of each polymerase for GC-based
amplification differences, sequence error rate, number of sequences with
an error, base substitution, and numbers of chimeras before and after
chimera removal with VSEARCH was assessed. Additionally, the correlation
between the number of chimeras and the number of OTUs was also assessed.
The total number of OTUs, taxa differences, and Bray-Curtis indices were
analyzed for both the fecal and mock community samples. Finally, Random
Forest models were created to assess whether classification of
polymerase, cycles, or individual could be performed best using 16S rRNA
gene sequencing data. Additionally, overlap between the most important
OTUs to the three models was assessed using mean decrease in accuracy
(MDA).

\textbf{\emph{Statistical analysis.}} All analysis was done with the R
(v 3.4.4) software package (R Core Team, 2017). Data transformation and
graphing was completed using the tidyverse package (v 1.2.1) and colors
selected using the viridis package (v 0.4.1) (Garnier, 2017; Wickham,
2017). High and low GC content was determined based on the median GC
percentage of either the V4 region or the whole genome of the bacterial
species used in the mock community. Differences in the total number of
OTUs were analyzed using an ANOVA with a tukey post-hoc test. For the
comparison of the number of OTUS in fecal samples the data was
normalized to each individual by cycle number to account for the
biological variation. Bray-Curtis distance matrices were generated using
mothur. The distance matrix data was analyzed using PERMANOVA with the
vegan package (v 2.4.5) (Oksanen et al., 2017) and Kruskal-Wallis tests
within R. The Random Forest models were run using the caret package (v
6.0.78) (Jed Wing et al., 2017). A total of 100 10-fold CV runs on
different 80/20 splits of the data was run to generate a range of the
Logloss value. The probability of a correct call was obtained from this
Logloss value by taking the negative natural logrithim. For both error
and chimera analysis, samples were tested using Kruskal-Wallis with a
Dunns post-hoc test. Where applicable, correction for multiple
comparison utilized the Benjamini-Hochberg method (Benjamini \&
Hochberg, 1995).

\textbf{\emph{Reproducible methods.}} The code and analysis can be found
here \url{https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017}.
The raw sequences can be found on the SRA at the following accession
number SRP132931.

@@

\textbf{\emph{Observed relative abundance differs by GC content.}} There
was a significant difference in relative abundance between high/low GC
content based on either the V4 16S rRNA gene region or the whole genome
{[}Figure 1 \& Table S1{]}.

\textbf{\emph{Sequence error varies by polymerase and is consistent
across the number of cycles used.}} The median error rate varied by
polymerase, with Kappa having the highest error rate of all the
polymerases across the number of cycles {[}Figure 4 \& Table S3{]}. The
majority of the differences across the number of cycles was between
Kappa and the other polymerases {[}Figure 4 \& Table S4{]}. The total
sequences with at least one error was also polymerase dependent, with
the majority of differences being between Kappa or Accuprime and the
other polymerases {[}Table S3 \& S4{]}. These differences in error rates
were not due to polymerase dependent differences in base substitution
rate {[}Figure S1{]}. Collectively, the results suggest that sequence
error is dependent on polymerase, persists across the number of cycles
used, and are not due to any bias towards a specific base substitution.
